---
title: "Reproducible Research: Peer Assessment 1"
output:
  html_document:
    keep_md: true
---
load the required libraries
```{r, warning=FALSE,message=FALSE}
library(tidyverse)
library(lubridate)
library(stats)
```

## Loading and preprocessing the data
```{r}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
temp <- tempfile()
download.file(url, temp)
data <- read.table(unz(temp, "activity.csv"), sep =",", header = TRUE )
unlink(temp)
data <- as_tibble(data)
data <- mutate(data,
                day_of_week = wday(date),
                weekday_weekend = if_else(day_of_week == 1 | day_of_week == 7, "weekend", "weekday")
               )
data$weekday_weekend <- as.factor(data$weekday_weekend)
data_group_1 <- group_by(data, date) %>% summarise(steps = sum(steps))
head(data)
```

## What is mean total number of steps taken per day?
Histogram of Daily Steps
```{r}
ggplot(data_group_1, aes(x=steps)) +
        geom_histogram(binwidth=500) +
        labs(x = "Steps per Day", y = "Frequency", title = "Daily Number of Steps")
```
```{r}
summary(data_group_1$steps)
```


## What is the average daily activity pattern?
This is an graph showing average daily activity at each interval
```{r}
data_group_2 <- group_by(data, interval) %>% summarise(steps = mean(steps, na.rm = TRUE))

ggplot(data=data_group_2, aes(x=interval, y=steps, group=1)) +
        geom_line() +
        labs(x = "5-minute Intervals", y = "Number of Steps", title = "Average Steps Taken")
```

This is the interval with the greatest average activity
```{r}
data_group_2[which.max(data_group_2$steps),]
```

## Imputing missing values
Find the number of missing values
```{r}
sum(is.na(data$steps))
```

I use a simple random imputation. For this approach I impute missing values based on the observed data for the variable.
```{r}
random.imp <- function (a){
        missing <- is.na(a)
        n.missing <- sum(missing)
        a.obs <- a[!missing]
        imputed <- a
        imputed[missing] <- sample (a.obs, n.missing, replace=TRUE)
        return (imputed)
}
df.steps <- random.imp(data$steps)
data$imputed = paste(random.imp(data$steps))
```

Histogram of Steps per Day
```{r}
data$imputed <- as.numeric(data$imputed)
data_group_3 <- group_by(data, date) %>% summarise(imputed = sum(imputed))
ggplot(data_group_3, aes(x=imputed)) +
        geom_histogram(binwidth=500) +
        labs(x = "Steps per Day", y = "Frequency", title = "Daily Number of Steps (imputed)")
```
```{r}
summary(data_group_3$imputed)
```
These values differ slightly. In the first part, the median was 10765 and the mean was 10766. Here the median is 10597 and the mean is 10657. Also, compared to the first set, the second set's 1st quartile has increased and the 3rd quartile has decreased. By imputing the missing values it helped to account for some outliers, and actually brought the both the mean and the median down.


## Are there differences in activity patterns between weekdays and weekends?
```{r}
ImputedStepsInterval <- data %>%
        group_by(interval,weekday_weekend) %>%
        summarise(imputed = mean(imputed))

ggplot(ImputedStepsInterval, aes(x=interval, y=imputed)) +
        geom_line() +
        facet_grid(weekday_weekend ~ .) +
        labs(x="5-Minute Intervals", y="Number of Steps", title = "Weekday vs. Weekend Activity (imputed)")
 ```

 During the weekday you see a high maximum number of steps followed by a period of low activity. This differs from the weekend where there appears to be a generally moderate level of sustained activity. Despite having the interval with the highest maximum, the weekday has a lower level of overall activity when compared to the weekend.
